---
title: "Private SNPS - Dmel Stealth Invasions"
output: rmarkdown::github_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Users/rpianezza/TE/')
```

In this script I will try to detect the private SNPs for each population that are driving the PCA. I want to confirm that the pattern shown by UMAP and PCA is confirmed directly from the data. To filter the SNPs, I used the script *mpileup2PCA.py* with low stringency (min-count 5, min-freq 0.1) to call many SNPs to work on in this script.

## Metadata and data preparation

We focus on **GDL** samples, the same used in the UMAP/PCA.
```{r}
library(tidyverse)

theme_set(theme_bw())

full_meta <- read_tsv("/Volumes/Temp1/Dmel-stealthTEs/preparation/dataset-metadata")

meta_gdl <- full_meta %>% select(run_accession, location, study, estimated_year) %>% rename(ID = "run_accession") %>% filter(study=="gdl") %>% select(-study, -estimated_year)
```

Upload the .PCAable files produced using the script *mpileup2PCA.py* and merging them with the metadata.
```{r}
te412 <- read_tsv("/Volumes/Temp1/Dmel-stealthTEs/private-SNPs/412/cleaned.PCAable")
te412$ID <- str_remove(te412$ID, "\\.cleaned$")
meta_412 <- inner_join(meta_gdl, te412, by="ID")

TIRANT <- read_tsv("/Volumes/Temp1/Dmel-stealthTEs/private-SNPs/tirant/cleaned.PCAable")
TIRANT$ID <- str_remove(TIRANT$ID, "\\.cleaned$")
meta_TIRANT <- inner_join(meta_gdl, TIRANT, by="ID")

BLOOD <- read_tsv("/Volumes/Temp1/Dmel-stealthTEs/private-SNPs/BLOOD/cleaned.PCAable")
BLOOD$ID <- str_remove(BLOOD$ID, "\\.cleaned$")
meta_BLOOD <- inner_join(meta_gdl, BLOOD, by="ID")

OPUS <- read_tsv("/Volumes/Temp1/Dmel-stealthTEs/private-SNPs/OPUS/cleaned.PCAable")
OPUS$ID <- str_remove(OPUS$ID, "\\.cleaned$")
meta_OPUS <- inner_join(meta_gdl, OPUS, by="ID")

CIRC <- read_tsv("/Volumes/Temp1/Dmel-stealthTEs/private-SNPs/CIRC/cleaned.PCAable")
CIRC$ID <- str_remove(CIRC$ID, "\\.cleaned$")
meta_CIRC <- inner_join(meta_gdl, CIRC, by="ID")

INVADER4 <- read_tsv("/Volumes/Temp1/Dmel-stealthTEs/private-SNPs/INVADER4/cleaned.PCAable")
INVADER4$ID <- str_remove(INVADER4$ID, "\\.cleaned$")
meta_INVADER4 <- inner_join(meta_gdl, INVADER4, by="ID")

I <- read_tsv("/Volumes/Temp1/Dmel-stealthTEs/private-SNPs/I/cleaned.PCAable")
I$ID <- str_remove(I$ID, "\\.cleaned$")
(meta_I <- inner_join(meta_gdl, I, by="ID"))

P <- read_tsv("/Volumes/Temp1/Dmel-stealthTEs/private-SNPs/P-element/cleaned.PCAable")
P$ID <- str_remove(P$ID, "\\.cleaned$")
(meta_P <- inner_join(meta_gdl, P, by="ID"))

hobo <- read_tsv("/Volumes/Temp1/Dmel-stealthTEs/private-SNPs/hobo/cleaned.PCAable")
hobo$ID <- str_remove(hobo$ID, "\\.cleaned$")
(meta_hobo <- inner_join(meta_gdl, hobo, by="ID"))
```

## Functions

Function which finds private SNPs in each population. Takes as input a merged file (PCAable + metadata) and a frequency threshold (es. 0.5). It counts how many SNPs are present in at least 50% of the individuals in a population and not present in more than 15 other samples.

```{r}
private <- function(data, threshold, percentage_pop_with_snp){
snps <- colnames(data)[3:length(colnames(data))]
populations <- data %>% select(location) %>% distinct() %>% pull()
private_snps <- tibble(location=NULL, SNP=NULL, thr=NULL, perc_shared_in_pop=NULL, number_external=NULL)

for (pop in populations){
  for (snp in snps){
    p <- data %>% select(ID, location, !!snp) %>% filter(location==pop)
    other <- filter(data, location!=pop)
    shared <- sum(p[[snp]] < threshold)
    count_pop <- p %>% summarise(count = n()) %>% pull()
    other_shared <- sum(other[[snp]] < threshold)
    if ((shared > (count_pop/100*percentage_pop_with_snp)) & other_shared<15){
      row <- tibble(location=pop, SNP=snp, thr=threshold, perc_shared_in_pop=shared/count_pop, number_external=other_shared)
      private_snps <- bind_rows(private_snps, row)
    }
  }
}
private_snps
}
```

Function to loop the function "private()" over multiple thresholds (from 0.5 to 0.9). A private SNP with 0.5 frequency is a very strong proof of a bottleneck during the invasion, while a SNP with only 0.9 is not that significant.

```{r}
different_thresholds <- function(data, percentage_pop_with_snp){
result_tibble <- tibble(location=NULL, SNP=NULL, thr=NULL, perc_shared_in_pop=NULL, number_external=NULL)
for (t in seq(0.5, 0.9, +0.1)){
  tib <- private(data, t, percentage_pop_with_snp)
  result_tibble <- bind_rows(result_tibble, tib)
}
pops <- result_tibble %>% select(location) %>% distinct() %>% pull()
filtered_tibble <- tibble(location=NULL, SNP=NULL, thr=NULL, perc_shared_in_pop=NULL, number_external=NULL)
for (pop in pops){
  only_pop <- filter(result_tibble, location==pop)
  filtered <- only_pop %>% filter(thr==0.5)
for (t in seq(0.6, 0.9, +0.1)){
  th <- filter(only_pop, thr<t) %>% select(SNP) %>% pull()
  non_overlapping <- filter(only_pop, thr==t, !(SNP %in% th))
  filtered <- bind_rows(filtered, non_overlapping)
}
  filtered_tibble <- bind_rows(filtered_tibble, filtered)
}
filtered_tibble
}
```

Function which takes as input the tibble returned from "different_thresholds" and creates a barplot.

```{r}
private_plot <- function(data, titlee){
  plottable <- data %>% group_by(location, thr) %>% summarise(count = n()) %>% mutate(thr=1-thr)
  ggplot(plottable, aes(x = thr, y = count, fill = location)) +
  geom_bar(stat = "identity") +
  labs(x = "Frequency threshold", y = "Private SNPs", fill = "Location") +
  ggtitle(titlee) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  ylim(0,25) #+ xlim(0,0.6)
}
```

## Plots

Create the tibbles of private alleles with different frequency thresholds for all the 4 interesting TEs.
```{r}
(tib412 <- different_thresholds(meta_412, 50))
(tib_BLOOD <- different_thresholds(meta_BLOOD, 50))
(tib_OPUS <- different_thresholds(meta_OPUS, 50))
(tib_TIRANT <- different_thresholds(meta_TIRANT, 50))
(tib_I <- different_thresholds(meta_I, 50))
#(tib_P <- different_thresholds(meta_P, 30))
(tib_hobo <- different_thresholds(meta_hobo, 50))
#tib_CIRC <- different_thresholds(meta_CIRC, 50)
#tib_INVADER4 <- different_thresholds(meta_INVADER4, 50)
```

Plot the data.
```{r}
(plot412 <- private_plot(tib412, "412"))
(plotOPUS <- private_plot(tib_OPUS, "Opus"))
(plotTIRANT <- private_plot(tib_TIRANT, "Tirant"))
(plotI <- private_plot(tib_I, "I-element"))
#(plotP <- private_plot(tib_P, "P-element"))
#(plotCIRC <- private_plot(tib_CIRC, "CIRC"))
#(plotINVADER4 <- private_plot(tib_INVADER4, "INVADER4"))

#ggsave("/Volumes/Temp1/Dmel-stealthTEs/private-SNPs/412.png", plot412, dpi = 300, width = 8, height = 6)
#ggsave("/Volumes/Temp1/Dmel-stealthTEs/private-SNPs/OPUS.png", plotOPUS, dpi = 300, width = 8, height = 6)
#ggsave("/Volumes/Temp1/Dmel-stealthTEs/private-SNPs/TIRANT.png", plotTIRANT, dpi = 300, width = 8, height = 6)
#ggsave("/Volumes/Temp1/Dmel-stealthTEs/private-SNPs/CIRC.png", plotCIRC, dpi = 300, width = 8, height = 6)
```

```{r}
private_plot_BLOOD<- function(data, titlee){
  plottable <- data %>% group_by(location, thr) %>% summarise(count = n()) %>% mutate(thr=1-thr)
  add <- tibble(location=c("Netherlands", "Tasmania, Australia", "Ithaca, USA"), thr=c(0.3, 0.4, 0.5), count = 0)
  plottable <- bind_rows(plottable, add)
  ggplot(plottable, aes(x = thr, y = count, fill = location)) +
  geom_bar(stat = "identity") +
  labs(x = "Frequency threshold", y = "Private SNPs", fill = "Location") +
  ggtitle(titlee) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  ylim(0,25) #+ xlim(0,0.6)
}

(plotBLOOD <- private_plot_BLOOD(tib_BLOOD, "Blood"))
ggsave("/Volumes/Temp1/Dmel-stealthTEs/private-SNPs/BLOOD.png", plotBLOOD, dpi = 300, width = 8, height = 6)

private_plot_HOBO<- function(data, titlee){
  plottable <- data %>% group_by(location, thr) %>% summarise(count = n()) %>% mutate(thr=1-thr)
  add <- tibble(location=c("Tasmania, Australia", "Zimbabwe"), thr=0.1, count = 0)
  plottable <- bind_rows(plottable, add)
  ggplot(plottable, aes(x = thr, y = count, fill = location)) +
  geom_bar(stat = "identity") +
  labs(x = "Frequency threshold", y = "Private SNPs", fill = "Location") +
  ggtitle(titlee) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  ylim(0,25) #+ xlim(0,0.6)
}
(plothobo <- private_plot_HOBO(tib_hobo, "Hobo"))
```