---
title: "Dataset preparation"
output: rmarkdown::github_document
editor_options: 
  markdown: 
    wrap: sentence
    author: Riccardo & Almorò
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

The dataset we are going to use for our analysis is composed by:

* Global diversity lines (**gdl**) - 86 samples
* Old lab strains from the Tirant paper (**tirant**, ols) - 14 samples
* Museum specimens from 1800 (**museum**) - 25 samples

In total, we have **125 samples**.

The **metadata** file preparation is described in **metadata.md**.
Here, I want to describe the bioinformatic steps that were followed to prepare the dataset.


## Data download

The files were downloaded in *FASTQ.gz* format from the ENA website using **wget**.
The accession numbers are:

* PRJNA268111 (**gdl**)
* PRJNA634847 (**tirant**)
* PRJNA945389 (**museum**)

Example of the wget command, with the file *ftp.txt* containing all the ftp addresses for the files, one in each line.
```
wget -i ftp.txt -P /Volumes/INTENSO/data-meseum_invasions/old-lab-strains/
```

After downloading the files, we checked the **md5 sum** of all the files to assess their completeness. Example md5 for all the fastq.gz files in the folder:
```
md5 *fastq.gz > md5-downloaded.txt
```

Both the ftp and the expected md5 can be found in the ENA website report for the project (download in tsv format after selecting the columns of interest).


## Trimming

To be consistent with further analysis across all the samples, we decided to **trim** the reads of all the files to a length of **100 bp**. This threshold was decided after careful analysis of the read length distribution in all the samples.

Before starting the trimming we measured the length distribution of every sample with the following command:
```
for FILE in *; do gzip -cd $FILE|awk '{print $1}'|paste - - - -|awk '{print length($2)}'|sort|uniq -c > $FILE.txt; done
```


Then we trimmed all the reads to 100nt and we discarded all the reads under 100nt:
```
ls *gz | parallel -j 4 'gzip -cd {} | cut -c-100 | awk "NR%4==2 && length(\$0)>=100{print prev; print \$0; getline; print \$0; getline; print \$0} {prev=\$0}" | gzip -c > trimmed/{}'
```
[GNU Parallel](https://doi.org/10.5281/zenodo.7761866) was used to parallelize the process, making it faster.


After trimming we controlled the read length distribution of the trimmed files using the previous command, this time with parallelization to speed it up, the expected output of this control is to have all the reads at 100nt, with the total number of reads in the trimmed files equal to the sum of all the reads with more than 100nt in the original files.
```
ls * | parallel --jobs 10 'gzip -cd {} | awk "{print \$1}" | paste - - - - | awk "{print length(\$2)}" | sort | uniq -c > length_post_trimming/museum_post_trimming_length/{}.txt'
```


The two trimmed fastq files for sample are then merge because deviaTE analysis only takes single reads as input, to merge the two
files we used the following command:
```
find . -name '*_1.fastq.gz' | parallel -j4 'n={}; n=${n%_1.fastq.gz}; gzip -cd {} ${n}_2.fastq.gz | gzip -c > merged/${n}.fastq.gz'
```


## Mapping to reference library (bam files)

After trimming, the next step was to map the fastq files to the TEs reference library (*/ref/teseqs-3scg-dmel.fasta*).

Command for mapping:
```
bwa bwasw
```
