---
title: "Metadata preparation"
output: rmarkdown::github_document
editor_options: 
  markdown: 
    wrap: sentence
    author: Riccardo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Users/rpianezza/TE/')
```

```{r}
library(tidyverse)
```

I created a **unique metadata file** for the three merged datasets we are going to use:

* Global diversity lines (**gdl**) - 86 samples
* Old lab strains from the Tirant paper (**tirant**, ols) - 12 samples (3 files are RNA seq)
* Museum specimens from 1800 (**museum**) - 25 samples

The file contains the following information:

* **Sample identifier** (should match the name of the downloaded files)
* **Sample name**
* **Year** of collection
* **Location** of collection
* **Latitude** of collection
* **Longitude** of collection
* **Source study**

In total, we have **123 samples**. Note that for some samples we have some missing data (location, year).

```{r}
gdl <- read_tsv("/Volumes/Temp1/Dmel-stealthTEs/gdl/metadata-gdl.txt") %>% select(-sample_accession)
ols <- read_tsv("/Volumes/Temp1/Dmel-stealthTEs/old-lab-strains/metadata.txt") %>% filter(!(run_accession %in% c("SRR11846566","SRR11846567","SRR12831808")))
museum <- read_tsv("/Volumes/Temp1/Dmel-stealthTEs/museum/metadata-museum.txt") %>% select(-sample_accession)
```

```{r}
museum2 <- read_tsv("/Volumes/Temp1/Dmel-stealthTEs/museum/metadata-museum2.txt") %>% select(-attribution, -sra_accession)
museum_final <- museum %>% inner_join(museum2, by="sample") %>% mutate(study="museum")
```

```{r}
gdl_final <- gdl %>% 
  mutate(location = case_when(startsWith(sample, "B") ~ "Beijing, China", startsWith(sample, "I") ~ "Ithaca, USA", startsWith(sample, "N") ~ "Netherlands", startsWith(sample, "T") ~ "Tasmania, Australia", startsWith(sample, "Z") ~ "Zimbabwe"), lat = case_when(startsWith(sample, "B") ~ 40, startsWith(sample, "I") ~ 42, startsWith(sample, "N") ~ 52, startsWith(sample, "T") ~ 43, startsWith(sample, "Z") ~ 19), , long = case_when(startsWith(sample, "B") ~ 116, startsWith(sample, "I") ~ 76, startsWith(sample, "N") ~ 5, startsWith(sample, "T") ~ 147, startsWith(sample, "Z") ~ 29), study="gdl")
```

```{r}
metadata <- bind_rows(museum_final, ols, gdl_final)
metadata

write_tsv(metadata, "/Volumes/Temp1/Dmel-stealthTEs/dataset-metadata")
```

